---
title: "Filed Under: birth timing"
output: html_notebook
---

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(data.table)
library(rstatix)
library(scales)
```


## Data

CDC WONDER, National Vital Statistics System — Natality, 2016–2024 expanded.
Downloaded May 23, 2026. Years included: 2022, 2023, 2024.

Query parameters:
- Delivery method: Cesarean & Vaginal
- Group by: State of Residence; Year; Month; Weekday; Time of Day
- Zero values and suppressed values excluded

> Centers for Disease Control and Prevention, National Center for Health Statistics.
> National Vital Statistics System, Natality on CDC WONDER Online Database.
> Data are from the Natality Records 2016–2024, as compiled from data provided by
> the 57 vital statistics jurisdictions through the Vital Statistics Cooperative Program.
> Accessed at http://wonder.cdc.gov/natality-expanded-current.html on May 23, 2026.

Data were downloaded by year and delivery method and combined for analysis.

Variables:
- `state_of_residence`
- `year` — 2022, 2023, or 2024
- `month`
- `weekday` — Weekday or Weekend
- `time_of_day` — binned into 3-hour blocks
- `delivery_method` — Cesarean or Vaginal
- `births` — count


```{r load}
time_of_day_data <- readr::read_csv("data/time_of_birth_dataset.csv",
                                    show_col_types = FALSE)

cat(sprintf("Dataset: %d observations, %d variables\n",
            nrow(time_of_day_data), ncol(time_of_day_data)))
```


## Completeness checks

For each combination of time of day, weekday, month, year, and delivery method —
how many of the 51 reporting areas (50 states + DC) contributed data?

```{r completeness}
time_of_day_data <- time_of_day_data %>%
  left_join(
    time_of_day_data %>%
      group_by(delivery_method, time_of_day, weekday, month, year) %>%
      count(name = "states_reported"),
    join_by(year, month, weekday, time_of_day, delivery_method)
  )

time_of_day_data %>%
  rstatix::get_summary_stats(states_reported, type = "quantile",
                             probs = c(0.1, 0.25, 0.5, 0.75, 0.9))
```

Analysis is limited to timepoints where at least 47 of 51 areas reported data
(≥ 90th percentile of states_reported), covering ~75% of raw observations.

```{r filter}
time_of_day_data_filtered <- time_of_day_data %>% filter(states_reported >= 47)

cat(sprintf("Filtered dataset: %.0f%% of raw observations retained\n",
            100 * nrow(time_of_day_data_filtered) / nrow(time_of_day_data)))
```

States with no weekend cesarean deliveries in the dataset:

```{r weekend-c-coverage}
weekend_c_states  <- time_of_day_data %>%
  filter(weekday == "Weekend", delivery_method == "cesarean") %>%
  distinct(state_of_residence) %>% pull()

weekend_all_states <- time_of_day_data %>%
  filter(weekday == "Weekend") %>%
  distinct(state_of_residence) %>% pull()

setdiff(weekend_all_states, weekend_c_states)
```


## Births by time of day

Average births per day (adjusted for 5 weekdays vs. 2 weekend days), by delivery
method and time of day.

```{r wrangle}
time_of_births_df <- time_of_day_data %>%
  select(state_of_residence, time_of_day, time_of_day_code,
         weekday, delivery_method, births) %>%
  group_by(time_of_day, time_of_day_code, state_of_residence,
           weekday, delivery_method) %>%
  summarize(total_births = sum(births, na.rm = TRUE), .groups = "drop") %>%
  mutate(births_per_day_ave = case_when(
    weekday == "Weekday" ~ total_births / 5,
    weekday == "Weekend" ~ total_births / 2
  ))
```

Medians by time of day, delivery method, and weekday/weekend:

```{r medians}
time_of_births_df %>%
  group_by(time_of_day, delivery_method, weekday) %>%
  summarise(
    median_births_per_day = median(births_per_day_ave, na.rm = TRUE),
    n_states              = n(),
    .groups = "drop"
  ) %>%
  arrange(delivery_method, weekday, time_of_day)

quantiles_df <- time_of_births_df %>%
  group_by(time_of_day_code, delivery_method, weekday) %>%
  rstatix::get_summary_stats(births_per_day_ave, type = "quantile",
                             probs = c(0.1, 0.25, 0.5, 0.90, 1))
```


## Statistical tests

Wilcoxon rank-sum tests (Mann–Whitney U) compare average births per day between
weekday and weekend groups for each time window and delivery method. Two-sided.
P-values adjusted for multiple comparisons using Benjamini–Hochberg (16 tests:
8 time windows × 2 delivery methods).

Shapiro–Wilk is included for transparency. Wilcoxon is used regardless, given
expected right-skew from state-level birth volume variation (California and
Wyoming are in the same pool).

```{r normality-check}
normality_check <- time_of_births_df %>%
  group_by(time_of_day, delivery_method, weekday) %>%
  shapiro_test(births_per_day_ave) %>%
  select(time_of_day, delivery_method, weekday, statistic, p) %>%
  arrange(delivery_method, time_of_day, weekday)

normality_check
```

```{r wilcoxon}
wilcox_results <- time_of_births_df %>%
  group_by(time_of_day, time_of_day_code, delivery_method) %>%
  wilcox_test(births_per_day_ave ~ weekday, paired = FALSE) %>%
  adjust_pvalue(method = "BH") %>%
  add_significance(p.col = "p.adj")

medians_wide <- time_of_births_df %>%
  group_by(time_of_day, time_of_day_code, delivery_method, weekday) %>%
  summarise(median_bpd = round(median(births_per_day_ave, na.rm = TRUE)),
            .groups = "drop") %>%
  pivot_wider(names_from = weekday, values_from = median_bpd)

results_table <- wilcox_results %>%
  left_join(medians_wide, by = c("time_of_day", "time_of_day_code", "delivery_method")) %>%
  arrange(delivery_method, time_of_day_code) %>%
  select(time_of_day, delivery_method, Weekday, Weekend, statistic, p, p.adj, p.adj.signif)

results_table
```


## Finding

Weekend deliveries — both vaginal and cesarean — are distributed roughly evenly
across the 24-hour day.

Weekday vaginal deliveries shift toward the afternoon. Weekday cesarean deliveries
shift toward the morning. The daytime clustering on weekdays is absent on weekends,
when scheduled procedures are uncommon.

Two states (Vermont and Wyoming) recorded no weekend cesarean deliveries in this
dataset across 2022–2024.


## Observations

Cesarean deliveries differ significantly between weekday and weekend in four
consecutive time windows spanning 06:00–17:59. The largest difference is in the
06:00–08:59 window: weekday median 1,818 births per day versus 503 on weekends
(W = 1843.0, p.adj < 0.001). The 09:00–11:59 (W = 1712.0, p.adj = 0.008),
12:00–14:59 (W = 1734.5, p.adj = 0.001), and 15:00–17:59 (W = 1612.5,
p.adj = 0.013) windows are also significant after Benjamini–Hochberg correction.

No significant weekday–weekend difference is detected for cesarean deliveries
outside the 06:00–17:59 window. Overnight (00:00–05:59), weekday and weekend
cesarean volumes are statistically indistinguishable (p.adj > 0.75).

Vaginal deliveries show no significant weekday–weekend difference in any of the
eight time windows after correction (all p.adj > 0.09). The visual impression of
a weekday afternoon shift does not reach statistical significance at the adjusted
threshold.

---

*Filed Under. Observed in the wild.*
